Determining the alternatives for scalar implicature

### 2016-01-29 Paper/Paper.tex
minor edits, spelling, naming, before sending to MF

\begin{abstract}
Succesful communication regularly requires listeners to make pragmatic
inferences - enrichments beyond the literal meaning of a speaker's
utterance. For example, when interpreting a sentence such as ``Alice ate
some of the cookies,'' listeners routinely infer that Alice did not eat
all of them. A Gricean account of this phenomena assumes the presence of
alternatives (like ``all of the cookies'') with varying degrees of
informativity, but it remains an open question precisely what these
alternatives are. We use a computational model of pragmatic inference to
test hypotheses about how well different sets of alternatives allow us
to predict scalar implicature performance across a range of different
scales. Our findings suggest that human comprehenders likely consider a
much broader set of alternatives beyond those entailed by the initial
description.

@@

Franke (2014) used an RSA-style model to assess what alternatives a
speaker would need to consider in order to produce the
typicality/felicity ratings reported by Degen \& Tanenhaus (2015) and
Tiel (2014). In order to do this, Franke (2014)'s model assigned weights
to a set of alternative numerical expressions. Surprisingly, along with
weighting \emph{one} highly (a conclusion that was suppored by the
empirical work), the best-fitting model assigned substantial weight to
\emph{none} as an alternative. This finding was especially surprising
considering the emphasis of standard theories on scalar items that stand
in entailment relationships with one another (e.g. \emph{one} entails
\emph{some} even if it is not classically considered to be part of the
scale).

@@

We begin by giving a brief presentation of the basic RSA model. This
model simulates the judgements of a pragmatic listener who wants to
infer a speaker's intended meaning \(m\) from her utterance \(u\). For
simplicity, we present a version of this model in which there is only
full recursion: that is, the pragmatic listener reasons about a
pragmatic speaker, who in turn reasons about a ``literal listener.'' We
assume throughout that this computation takes place in a signaling game
(Lewis, 1969) with a fixed set of possible meanings \(m \in M\) and a
fixed possible set of utterances \(u \in U\), with both known to both
participants. Our goal in this study is to determine what utterances
fall in \(U\).

@@

This model of the pragmatic speaker is consistent with a speaker who
chooses words to maximize the utility of an utterance in context (Frank
\& Goodman, 2012), where utility is operationalized as the informativity
of a particular utterance (surprisal) minus a cost:

@@

\noindent where \(C(u)\) is the cost of a particular utterance,
\(-log(p_{L_0})\) represents the \emph{surprisal} for the literal
listener (the information content of the utterance), and (\(\alpha\) is
a parameter in a standard choice rule. If \(\alpha=0\), speakers choose
randomly and as \(\alpha \rightarrow \infty\), they greedily choose the
highest probability alternative. In our simulations below, we treat
\(\alpha\) as a free parameter and fit it to the data.

To instantiate our signaling game with a tractable message set \(M\), in
our studies we adopt a food-review paradigm: we assume that speakers and
listeners are trying to communicate the number of stars in an online
restaurant review (where \(m \in \{1, 2, 3, 4, 5\}\)). We then use
experiments to measure three components of the model. First, to measure
literal semantics \({p_{L_0} (m \mid u)}\) (we ask experiment
participants to judge whether a message is applicable to a particular
meaning (Experiment 1). Second, to generate a set of plausible
alternative messages in \(U\), we elicit alternatives directly
(Experiment 2). Lastly, to obtain human \(L_1\) pragmatic judgments, we
ask participants to interpret a speaker's utterances (Experiment 3).

@@

Experiment 1 was conducted to approximate literal listener semantic
distributions \(p_{L_0}(m \mid u)\) for five pairs of scalar items taken
from Tiel (2014). We include three conditions in Experiment 1,
corresponding to the sets of alternatives within a scale that
participants were presented with: two alternatives (``entailment''),
four alternatives, and five alternatives. The two-alternative entailment
condition makes a test of the hypothesis that the two members of the
classic Horn scale (Horn, 1972) are the only alternatives necessary to
predict the strength of listeners' pragmatic inference. The four- and
five-alternative conditions then add successively more alternatives to
test whether including a larger number of alternatives will increase
model
fit.\footnote{Note that alternatives in the four- and five-alternatives conditions were chosen on the basis of Experiment 2, which was run chronologically after the two-alternative condition; all literal listener experiments are grouped together for simplicity in reporting.}
A secondary goal of Experiments 1b,c tests whether the total set of
alternatives queried during literal semantic elicitation effects
compatibility judgments. (If it does we should see differences in
compatibility judgments between experiments.) We address this possibiliy
in the results section.

@@

Figure \ref{fig:stimuli_exp1}, left, shows the experimental setup.
Participants were presented with a target scalar item and a star rating
(1--5 stars) and asked to judge the compatibility of the scalar item and
star rating. Compatibility was assessed through a binary ``yes/no''
response to a question of the form ``Do you think that the person
thought the food was \_\_\_\_?" where a target scalar was presented in
the blank. Each participant saw all scalar item and star rating
combinations for their particular condition, in a random order.

@@

\caption[Literal listener judgements from Experiments 1a,b,c]{Literal listener judgements from Experiments 1a,b,c. Proportion of participants indicating acceptability is shown on the vertical axis, with the horizontal axis showing number of stars on which the utterance was judged. Each line-type shows a different experiment, and colors indicate the different items in the scale (with the experiments including different numbers of items).}\label{fig:exp1Plots}
\end{figure*}
\end{CodeChunk}

Figure \ref{fig:exp2Plots} plots literal listener \(p_{L0}(m|u)\) scalar
item distributions for the three conditions. Each row plots a unique
scalar family with items ordered horizontally by valence. Several trends
are visible. First, in each scale, the alternatives spanned the star
scale, such that there were alternatives that were highly compatible
with both the lowest and highest numbers of stars. Second, there was
clear variability between scalar families. For example, compatibility
judgments for the top items in the \emph{memorable / unforgettable}
scale were more similar than those for \emph{good / excellent} or
\emph{liked / loved.} Finally, there was substantial consistency in
ratings for items that were repeated across experiments, suggesting that
this paradigm elicited stable judgments from participants.

To test our secondary hypothesis, that the different sets of scalar
items queried might result in differences in compatibility judgments
between experiments, we ran a mixed effects model. We regressed
compatibility judgments on scale, number of stars and experiment
(1a,b,c), with subject and word level random effects, which was the
maximal structure that converged. Results indicate no significant
differences between Experiment 1a and 1c \(b =\)-0.05, \(Z =\) -0.53,
\(p =\) 0.59 or between 1b and 1c \(b =\) -0.04, \(Z =\) -0.52 \(p =\)
0.6 and the addition of our experiment predictor did not significantly
improve model fit when compared to a model without the experiment
variable using ANOVA \(\chi^2(\) 2 \() =\) 0.43, \(p =\) 0.81.

@@

Participants were presented a target scalar term from our original
entailment set embedded in a sentence such as, ``In a recent restaurant
review someone said they thought they the food was \_\_\_\_," with a
target scalar presented in the blank. Participants were then asked to
generate plausible alternatives by responding to the question, ``If
they'd felt differently about the food, what other words could they have
used instead of \_\_\_\_?" They were prompted to generate three unique
alternatives.

@@

\caption[Combined counts for participant generated alternatives for 'liked  loved' scale]{Combined counts for participant generated alternatives for 'liked  loved' scale. Participants were shown a target scalar (either liked or loved) and asked to give three alternatives.}\label{fig:exp2_altsPlot_likedLoved}
\end{figure}
\end{CodeChunk}

@@

Experiment 3 was conducted to measure pragmatic judgments
\(p_{L_1}(m \mid u)\) for five pairs of scalar items taken from Tiel
(2014). We include two conditions in Experiment 3. In the
two-alternative condition, participants made judgments for items
included in the entailment scales. In the four-alternative condition,
participants made judgments for the entialment items and also the top
two alternatives elicited for each scale in Experiment 2. Running two
pragmatic judgment conditions allowed us to rule out the potential
effects of having a larger set of alternatives during the pragmatic
judgment elicitation and also provided two sets of human judgments to
compare with model predictions.
\footnote{Note that alternatives in the four-alternative condition were chosen on the basis of Experiment 2, which was run chronologically after the two-alternative condition; both pragmatic listener experiments are grouped together for simplicity in reporting.}
A secondary goal of Experiment 3b tests whether the total set of
alternatives queried in Experiments 3a,b effects pragmatic judgments.
(If it does we should see differences in pragmatic judgments between
experiments.) We address this possibiliy in the results section.

@@

Figure \ref{fig:exp2Plots} plots pragmatic listener judgments
distributions for ``weak'' / ``strong'' scalar pairs (ie
\emph{good}/\emph{excellent}). We only include the original entailment
pairs in this figure for simplicity. Several trends are visible. First,
in each scale participants generated implicatures and were less likely
to assign high star-ratings to weaker scalar terms, despite literal
semantic compatibility. Second, the ``size'' of the difference between
judgments between ``strong'' and ``weak'' scalar items varied by scale,
confirming results found by Tiel (2014).

To test our secondary hypothesis, that the different sets of scalar
items queried might result in differences in pragmatic judgments between
experiments, we ran a mixed effects model. We regressed pragmatic
judgements on scale and experiment (3a,b) with subject and word level
random effects, which was the maximal structure that converged. Results
indicate there were no significant differences between Experiment 3a and
3b for \(b =\) 0.05, \(t(\) 150 \() =\) 1.04, \(p =\) 0.3 and the
addition of our experiment predictor did not significantly improve model
fit when compared to a model without the experiment variable using ANOVA
\(\chi^2(\) 1 \() =\) 1.13, \(p =\) 0.29.

@@

By varying the type of scale representations available to our Bayesian
model we investigated the effects of alternatives on scalar implicature.
Model fit with human judgments was significantly improved by the
inclusion of alternatives beyond the typical ``strong'' and ``weak''
scalar items. The ``Two alternative'' model contained only entailment
items, which, under classic accounts, should be sufficient to generate
implicature. Model performance with this limited alternative set was
poor when compared to human judgments from Experiment 3a,b (row 1, Table
\ref{xtable}). Building off findings in Franke (2014), in which the
best-fitting model assigned substantial weight to \emph{none} as an
alternative, we included a ``generic'' negative alternative that was
only compatible with 1-star. Model fit for this modified ``Two
alternative with generic negative'' alternative set did improve, however
overall correlation was still low (row 2, Table \ref{xtable}). Model fit
jumped substantially by adding the top-two alternatives for each scale
(row 3, Table \ref{xtable}). Finally, our ``Five alternative'' model
included a neutral alternative along with the original entailment pair
and top two (row 4, Table \ref{xtable}).

While model improvement seemed to be related to the number of
alternatives present, we did not see substantial differences between our
four- and five-alternative models. One possibility is that alternatives
are differentially salient in context. Our current framework makes the
simplifying assumption that all alternatives are equally salient.
However, it may well be the case that while both \emph{all} and
\emph{most} are \emph{present} during pragmatic enrichment of
\emph{some}, \emph{all} is more salient and therefore should carry
greater weight in our model. Findings from Experiment 2, in which
participants generated plausible alternatives, supports this intuition,
as some alternatives appeared more often than others. This distinction
between binary \emph{presence} and graded \emph{salience} is an
important one that deserves further investigation.

It is also likely that the precise set of alternatives present during
implicature are domain dependent. Our current empirical paradigm
elicited literal semantics, pragmatic judgments and plausible
alternatives all within the restricted domain of food reviews.
Participant data may have differed greatly if we had instead asked them
to respond in a movie review paradigm, which would have been acceptable
for a number of our stimuli (ie \emph{memorable}/\emph{unforgettable},
\emph{liked}/\emph{loved}, etc.).

>>>>

\begin{abstract}
Successful communication regularly requires listeners to make pragmatic
inferences - enrichments beyond the literal meaning of a speaker's
utterance. For example, when interpreting a sentence such as ``Alice ate
some of the cookies,'' listeners routinely infer that Alice did not eat
all of them. A Gricean account of this phenomena assumes the presence of
alternatives (like ``all of the cookies'') with varying degrees of
informativity, but it remains an open question precisely what these
alternatives are. We use a computational model of pragmatic inference to
test hypotheses about how well different sets of alternatives allow us
to predict scalar implicature performance across a range of different
scales. Our findings suggest that human comprehenders likely consider a
much broader set of alternatives beyond those entailed by the initial
description.

@@

Franke (2014) used an RSA-style model to assess what alternatives a
speaker would need to consider in order to produce the
typicality/felicity ratings reported by Degen \& Tanenhaus (2015) and
Tiel (2014). In order to do this, Franke (2014)'s model assigned weights
to a set of alternative numerical expressions. Surprisingly, along with
weighting \emph{one} highly (a conclusion that was supported by the
empirical work), the best-fitting model assigned substantial weight to
\emph{none} as an alternative. This finding was especially surprising
considering the emphasis of standard theories on scalar items that stand
in entailment relationships with one another (e.g. \emph{one} entails
\emph{some} even if it is not classically considered to be part of the
scale).

@@

We begin by giving a brief presentation of the basic RSA model. This
model simulates the judgments of a pragmatic listener who wants to infer
a speaker's intended meaning \(m\) from her utterance \(u\). For
simplicity, we present a version of this model in which there is only
full recursion: that is, the pragmatic listener reasons about a
pragmatic speaker, who in turn reasons about a ``literal listener.'' We
assume throughout that this computation takes place in a signaling game
(Lewis, 1969) with a fixed set of possible meanings \(m \in M\) and a
fixed possible set of utterances \(u \in U\), with both known to both
participants. Our goal in this study is to determine what utterances
fall in \(U\).

@@

This model of the pragmatic speaker (denoted \(S_1\)) is consistent with
a speaker who chooses words to maximize the utility of an utterance in
context (Frank \& Goodman, 2012), where utility is operationalized as
the informativity of a particular utterance (surprisal) minus a cost:

@@

\noindent where \(C(u)\) is the cost of a particular utterance,
\(-log(p_{L_0})\) represents the \emph{surprisal} for the literal
listener (the information content of the utterance), and \(\alpha\) is a
parameter in a standard choice rule. If \(\alpha=0\), speakers choose
randomly and as \(\alpha \rightarrow \infty\), they greedily choose the
highest probability alternative. In our simulations below, we treat
\(\alpha\) as a free parameter and fit it to the data.

To instantiate our signaling game with a tractable message set \(M\), in
our studies we adopt a food-review paradigm: we assume that speakers and
listeners are trying to communicate the number of stars in an online
restaurant review (where \(m \in \{1, 2, 3, 4, 5\}\)). We then use
experiments to measure three components of the model. First, to measure
literal semantics \({p_{L_0} (m \mid u)}\) (we ask experiment
participants to judge whether a message is compatible with a particular
meaning (Experiment 1). Second, to generate a set of plausible
alternative messages in \(U\), we elicit alternatives directly
(Experiment 2). Lastly, to obtain human \(L_1\) pragmatic judgments, we
ask participants to interpret a speaker's utterances (Experiment 3).

@@

Experiment 1 was conducted to approximate literal listener semantic
distributions \(p_{L_0}(m \mid u)\) for five pairs of scalar items taken
from Tiel (2014). We include three conditions in Experiment 1,
corresponding to the sets of alternatives within a scale that
participants were presented with: two-alternatives (``entailment''),
four-alternatives, and five-alternatives. The two-alternatives
entailment condition makes a test of the hypothesis that the two members
of the classic Horn scale (Horn, 1972) are the only alternatives
necessary to predict the strength of listeners' pragmatic inference. The
four- and five-alternatives conditions then add successively more
alternatives to test whether including a larger number of alternatives
will increase model
fit.\footnote{Note that alternatives in the four- and five-alternatives conditions were chosen on the basis of Experiment 2, which was run chronologically after the two-alternative condition; all literal listener experiments are grouped together for simplicity in reporting.}
A secondary goal of Experiments 1a,b,c tests whether the set of
alternatives queried during literal semantic elicitation impacts
compatibility judgments. (If it does we should see differences in
compatibility judgments for shared items between experiments.) We
address this possibility in the results section.

@@

Figure \ref{fig:stimuli_exp1}, left, shows the experimental setup.
Participants were presented with a target scalar item and a star rating
(1--5 stars) and asked to judge the compatibility of the scalar item and
star rating. Compatibility was assessed through a binary ``yes/no''
response to a question of the form, ``Do you think that the person
thought the food was \_\_\_\_?" where a target scalar was presented in
the blank. Each participant saw all scalar item and star rating
combinations for their particular condition, in a random order.

@@

\caption[Literal listener judgments from Experiments 1a,b,c]{Literal listener judgments from Experiments 1a,b,c. Proportion of participants indicating compatibility (answering 'yes') is shown on the vertical axis, with the horizontal axis showing number of stars on which the utterance was judged. Rows are grouped by scale and items within rows are ordered by valence. Colors indicate the specific experiment (1a,b,c) with experiments including different numbers of items.}\label{fig:exp1Plots}
\end{figure*}
\end{CodeChunk}

Figure \ref{fig:exp2Plots} plots literal listener \(p_{L0}(m|u)\) scalar
item distributions for the three conditions. Each row shows a unique
scalar family with items ordered horizontally by valence. Several trends
are visible. First, in each scale, the alternatives spanned the star
scale, such that there were alternatives that were highly compatible
with both the lowest and highest numbers of stars. Second, there was
clear variability between scalar families. For example, compatibility
judgments for the top items in the \emph{memorable / unforgettable}
scale were more similar than those for \emph{good / excellent} or
\emph{liked / loved.} Finally, there was substantial consistency in
ratings for items that were repeated across experiments, suggesting that
this paradigm elicited stable judgments from participants.

To test our secondary hypothesis, that the different sets of scalar
items queried might result in differences in compatibility judgments
between experiments, we ran a mixed effects model. We regressed
compatibility judgments on scale, number of stars and experiment
(1a,b,c), with subject and word level random effects, which was the
maximal structure that converged. Results indicate no significant
differences between Experiment 1a and 1c, \(b =\)-0.05, \(Z =\) -0.53,
\(p =\) 0.59 or between 1b and 1c, \(b =\) -0.04, \(Z =\) -0.52 \(p =\)
0.6 and the addition of our experiment predictor did not significantly
improve model fit when compared to a model without the experiment
variable using ANOVA \(\chi^2(\) 2 \() =\) 0.43, \(p =\) 0.81.

@@

Participants were presented a target scalar item from our original
entailment set (see Figure \ref{fig:allScalesTable}) embedded in a
sentence such as, ``In a recent restaurant review someone said they
thought they the food was \_\_\_\_," with a target scalar presented in
the blank. Participants were then asked to generate plausible
alternatives by responding to the question, ``If they'd felt differently
about the food, what other words could they have used instead of
\_\_\_\_?" They were prompted to generate three unique alternatives.

@@

\caption[Combined counts for participant generated alternatives for 'liked  loved' scale from Experiment 2]{Combined counts for participant generated alternatives for 'liked  loved' scale from Experiment 2. Participants were shown a target scalar (either liked or loved) and asked to give three alternatives.}\label{fig:exp2_altsPlot_likedLoved}
\end{figure}
\end{CodeChunk}

@@

Experiment 3 was conducted to measure pragmatic judgments
\(p_{L_1}(m \mid u)\) for five pairs of scalar items taken from Tiel
(2014). We include two conditions in Experiment 3. In the
two-alternatives condition, participants made judgments for items
included in the entailment scales. In the four-alternatives condition,
participants made judgments for the entailment items and also the top
two alternatives elicited for each scale in Experiment 2. Running two
pragmatic judgment conditions allowed us to rule out the potential
effects of having a larger set of alternatives during the pragmatic
judgment elicitation and also provided two sets of human judgments to
compare with model
predictions\footnote{Note that alternatives in the four-alternatives condition were chosen on the basis of Experiment 2, which was run chronologically after the two-alternatives condition; both pragmatic listener experiments are grouped together for simplicity in reporting.}.
A secondary goal of Experiment 3b tests whether the set of alternatives
queried in Experiments 3a,b impacts pragmatic judgments. (If it does we
should see differences in pragmatic judgments between experiments.) We
address this possibility in the results section.

@@

Figure \ref{fig:exp2Plots} plots pragmatic listener judgments
distributions for ``weak'' / ``strong'' scalar pairs (e.g.
\emph{good}/\emph{excellent}). We only include the original entailment
pairs in this figure for simplicity. Several trends are visible. First,
in each scale participants generated implicatures and were less likely
to assign high star-ratings to weaker scalar terms, despite literal
semantic compatibility. Second, the ``size'' of the difference between
judgments between ``strong'' and ``weak'' scalar items varied by scale,
consistent with results from Tiel (2014).

To test our secondary hypothesis, that the different sets of scalar
items queried might result in differences in pragmatic judgments between
experiments, we ran a mixed effects model. We regressed pragmatic
judgments on scale and experiment (3a,b) with subject and word level
random effects, which was the maximal structure that converged. Results
indicate there were no significant differences between Experiment 3a and
3b, \(b =\) 0.05, \(t(\) 150 \() =\) 1.04, \(p =\) 0.3 and the addition
of our experiment predictor did not significantly improve model fit when
compared to a model without the experiment variable using ANOVA
\(\chi^2(\) 1 \() =\) 1.13, \(p =\) 0.29.

@@

By varying the type of scale representations available to our Bayesian
model we investigated the effects of alternatives on scalar implicature.
Model fit with human judgments was significantly improved by the
inclusion of alternatives beyond the typical ``strong'' and ``weak''
scalar items. The two-alternatives model contained only entailment
items, which, under classic accounts, should be sufficient to generate
implicature. Model performance with this limited alternative set was
poor when compared to human judgments from Experiment 3a,b (row 1, Table
\ref{xtable}). Building off findings in Franke (2014), in which the
best-fitting model assigned substantial weight to \emph{none} as an
alternative, we included a ``generic'' negative alternative that was
only compatible with 1-star. Model fit for this modified entailment
alternative set did improve, however overall correlation was still low
(row 2, Table \ref{xtable}). Model fit jumped substantially by adding
the top-two alternatives for each scale (row 3, Table \ref{xtable}).
Finally, our five-alternatives model included a neutral alternative
along with the original entailment pair and top two (row 4, Table
\ref{xtable}).

While model improvement seemed to be related to the number of
alternatives present, we did not see substantial differences between our
four- and five-alternative models. One possibility is that alternatives
are differentially salient in context. Our current framework makes the
simplifying assumption that all alternatives are equally salient.
However, it may well be the case that while both \emph{all} and
\emph{most} are \emph{present} during pragmatic enrichment of
\emph{some}, \emph{all} is more salient and therefore should carry
greater weight in our model. Findings from Experiment 2, in which
participants generated plausible alternatives, support this intuition,
as some alternatives appeared more often than others. This distinction
between binary \emph{presence} and graded \emph{salience} is an
important one that deserves further investigation.

It is also likely that the precise set of alternatives present during
implicature are domain dependent. Our current empirical paradigm
elicited literal semantics, pragmatic judgments and plausible
alternatives all within the restricted domain of restaurant reviews.
Participant data may have differed greatly if we had instead asked them
to respond in a movie review paradigm, which would have been acceptable
for a number of our stimuli (e.g. \emph{memorable}/\emph{unforgettable},
\emph{liked}/\emph{loved}, etc.).
